Translation of B Implementations to the LLVM: Specification

### 2014-01-08 DOCUMENTATION/specification.tex
Added a file containing the specification of the code generation. It synthesizes information provided elsewhere before (translation.tex).

\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage{amssymb}
\usepackage[fleqn]{amsmath}
\usepackage[colorlinks=false]{hyperref}
\usepackage{xspace,framed,here,keystroke}
\usepackage{graphicx}
\pagestyle{headings}
\title{Translation of B Implementations to the LLVM: Control and Modularity}
\author{David Déharbe \and Valério Medeiros Jr}
\institute{Program of Graduate Studies in Systems and Computing (PPgSC) \\
  Federal University of Rio Grande do Norte \\
  Natal (Brazil)}
\date{May 2013}

@@

\begin{abstract}
  The aim of this work is to lay the foundations of a multi-platform code
  generator for the B method. In particular, this paper presents a translation
  procedure from a large subset of the B language for implementations towards
  LLVM source code. This translation is defined formally as a set of rules
  defined recursively on the abstract syntax for B implementations. It handles
  the following elements of the B language: simple data types, imperative
  instructions and component compositions.
\end{abstract}

\section{Introduction}

B is a formal refinement-based software design method~\cite{Abrial1996}. It has
a single language encompassing abstract constructs suitable for specification,
as well as classic imperative constructs for computer programming. The starting
point of a B development is a specification, called a \emph{machine}, that is
incrementally refined to an \emph{implementation}, where only imperative-like
constructs may be employed~\cite{Clearsy}. Such implementation may then be
translated to source code in a programming language, say C~\cite{ComenC} or
Ada. All steps in the B method are formally verified using certified theorem
proving technologies. However the translation to some programming language and
its subsequent compilation to the target platform do not benefit from the same
mathematical rigor. In practice, redundancy in the tool chains and execution
platform are employed to increase the level of confidence to the desired levels.

The goal of this work is to connect the B method to the LLVM compilation
framework~\cite{Lattner04LLVM}.  LLVM is a free, open-source project upon which
many compiling contemporary technologies and tools are based. LLVM provides an
intermediate assembly language upon which may be applied techniques such as
optimization, static analysis, code generation, debugging, etc. Providing a
connection from the B method to this framework gives the users of the B method
access to these technologies. Our approach is to define a translation from B0
(the subset of the B language that is used to describe imperative programs) to
the LLVM intermediate format.

Considering the correctness of the proposed translation, we shall rely upon
human inspection for the time being. The standard uniform semantic framework for
B and LLVM that would be necessary to carry out a proof of this correctness does
not exist (yet). A possible approach to mitigate the risk of having an error in
the translation would be to include assertions in the target code. Indeed the
artifacts produced in the B development have many such assertions and it would
be possible to translate at least some of them (B has an assertion instruction
that introduces additional verification conditions in the development). The
produced code could then be (at least partially) verified by testing the absence
of assertion violations.

\paragraph{Overview.} Following this introduction, the paper is organized as
follows. The implementation constructs of the B language, and selected elements
of the LLVM are presented in section~\ref{sec:b-impl}
and~\ref{sec:llvm}. Section~\ref{sec:overview} presents a bird's-eye view that
is then detailed in sections~\ref{sec:module} (for component modularity)
~\ref{sec:data} (for data aspects),~\ref{sec:expr} (expressions) and
instructions~\ref{sec:control}. We conclude and discuss future work
in~\ref{sec:conclusion}.

\section{Implementation Language in the B-method}
\label{sec:b-impl}

This section describes the B implementation constructs that are considered in
the compilation to the LLVM. This description introduces the notations used in
the definition of the translation. The elements of the B language are denoted
using a \B{sans serif} font.

Figure~\ref{tab:node-attr} summarizes the structure
of the abstract syntax that is relevant in the definition of the proposed
translation and introduces identifiers used in the definition of the
translation. For instance, and for the purpose of the translation, the relevant
components of a B implementation are: \B{id}, the name identifying the
implementation within the project; \B{see}, a set of references to other modules
containing definitions used in the implementation; \B{const}, a set of constant
names, their type and their values; \B{impo}, a sequence of optionally
prefixed B machines corresponding to external modules instantiated in the
implementation; \B{var}, a set of variable names, their type and possibly
additional functional restrictions forming the rest of the state; \B{init}, a
sequence of instructions executed upon initialization of the implementation;
\B{op}, a set of operations, each being an algorithmic description of the
different functionalities provided by the implementation.\DD{Include the
possibility that the import is a machine without implementation. Instead it
has a list of typed, implementable, variables.}

@@

\begin{figure}[H]

@@

LLVM aims to be a general, language agnostic, intermediate representation for
compilers. The compiler front-end translates the source programming language to
LLVM source code, upon which optimization and other static analysis tasks may be
performed, and the compiler back-end translates to the target platform assembly
language. One key feature of LLVM is that it is a single-static assignment (SSA)
language: a variable may only be assigned once. So for each assignment to a
variable found in the source language, the front-end shall generate a new
temporary variable in the LLVM program. For illustration,
figure~\ref{fig:ex-llvm} presents a LLVM program (for clarity, the corresponding
C code appears on the left) with two temporary variables \llvm{\%0} and
\llvm{\%1}.
\begin{figure}
\vspace*{-3em}
  \begin{minipage}[t]{.3\textwidth}
    \begin{tabbing}
      foo \= \kill  \\
      void inc(int * pi) \{\\
      \> *pi += 1; \\
      \}\end{tabbing}
  \end{minipage}
  \hfill
  \begin{minipage}[t]{.6\textwidth}
    \begin{tabbing}
      foo \= \kill  \\
      \llvm{define void @inc(i32* \%pi) nounwind \{} \\
      \llvm{entry:}                      \\
      \> \llvm{\%0 = load i32* \%pi}     \\
      \> \llvm{\%1 = add i32 \%0, 1}     \\
      \> \llvm{store i32 \%1, i32* \%pi} \\
      \> \llvm{ret void} \\
      \llvm{\}}
    \end{tabbing}
  \end{minipage}
  \caption{Simple example of a C function and its corresponding LLVM function.
    The first line contains the signature: return type \llvm{void}, the name
    \llvm{@inc} and one parameter named \llvm{\%pi} and typed \llvm{i32*}.  Next
    is the body with a single block, labeled \llvm{entry}, and temporary
    variables \llvm{\%0} and \llvm{\%1}, created in the conversion to SSA. The
    block has four instructions: \llvm{load}, \llvm{add}, \llvm{store} and
    \llvm{ret}. For instance, \llvm{\%1 = add i32 \%0, 1} is an addition
    (\llvm{add}), has result type \llvm{i32} and assigns to
    \llvm{\%1} the sum of variable \llvm{\%0} and integer
    literal \llvm{1}.}
  \label{fig:ex-llvm}
\end{figure}

The introduction to the LLVM that follows is focused in the features
that are relevant for the representation of B
implementations. Figure~\ref{fig:llvm-grammar} presents a grammar of this subset
of the LLVM.

@@

LLVM programs are organized into modules, one per translation unit. A module may
contain declarations of external entities (functions, constants) and definitions
of internal items (functions, variables, constants). All data is explicitly
typed. The name and type of external entities must be declared.  All names,
e.g. non-reserved identifiers, shall start with \llvm{@}, when they are global,
or\llvm{\%}, when they are local. For instance, \llvm{@max = external constant
  i32} declares \llvm{@max} as a 32-bit integer constant and \llvm{declare void
  @inc(i32*)} declares \llvm{@inc} as a function having as parameter a pointer
to an integer and \llvm{void} as a return type.

The type system contains the empty type \llvm{void}, an infinite, countable
number of integer types, one for each possible bit width (e.g. \llvm{i8} is the
type for 8-bit integers), and type constructors pointer (monadic \llvm{$\cdot$
  *}) and structure (polyadic \llvm{\{$\cdots$\}}). For instance \llvm{\{ i8*,
  i8, i8 \}} is the type for structures with three fields, the first having as
type pointer to \llvm{i8}. Grammar rule \nt{type\_def} states how types are
named, e.g.: \llvm{\%T1 = type \{i32, i32\}} and \llvm{\%T2 = type \{\%T1 *,
  \%T1 *\}}. Note that there is no \llvm{void *} type in the LLVM, instead
pointer values are of integer types.

Local entities are constants, variables or functions. An example of constant
definition is \llvm{@secret = constant i32 42}, it is composed of a name, type
and value. The elements of a variable definition are the name, type and code
generation attributes, e.g \llvm{@count = common global i32 zeroinitializer}.
Attributes \llvm{common}, \llvm{global} and \llvm{zeroinitializer} provide
information for target code generation: linkage type (the variable is merged
with other of the same name), the scope (global) and initialization of the
variable (zeroed is mandatory for common linkage). LLVM defines other attributes
for these aspects, but they are not needed in this work and their presentation
is omitted. For each such definition, a memory block is statically allocated to
stores the variable value. Function definitions are composed of the signature
and body. The signature contains the return type, name, parameters, and
attributes for target code generation. Attribute \llvm{nounwind} states that the
function never generate exceptions. The body is a sequence of blocks of
instructions in single-static assignment (SSA) form, i.e. are such that each
variable may be assigned exactly once.

The grammar rule \nt{inst} describes the different kind of instructions. All
instructions producing a value assign it to a fresh variable.  Instruction
\llvm{alloca} allocates a memory block the size of the given type on the stack
segment. This memory is automatically freed when the current frame is popped
from the stack. The LLVM does not include instructions for allocating from the
heap. Arithmetic operations are binary, and include signed quotient and
reminder. Arithmetic comparisons compare two values according to the first
operand and return a 1-bit integer value.  Instruction \llvm{call} invokes the
given function with the given arguments and assigns the result to a fresh
variable, which is of the given type.  In general, \llvm{getelementptr} gets the
address of an element in an aggregate object through indexing. In the target
LLVM subset, aggregate objects are structures, the first index has type
\llvm{i32} and value \llvm{0} to select the first structure at the given
location \nt{exp}, and the second index selects a field in that structure.
Instruction \llvm{load} assigns to a fresh variable \nt{name} the contents of a
memory address of type \nt{type} specified by \nt{exp} (e.g. in
fig.~\ref{fig:ex-llvm}).  Instruction \llvm{store} writes a value to memory
address (again, see fig.~\ref{fig:ex-llvm}).  Instruction \llvm{br} has two
forms.  In the first form, it directs conditionally the control flow to one of
two blocks.  If the given expression evaluates to \llvm{1}, control goes to the
first block, otherwise to the second. In the second form, the control jumps
unconditionally to the given block.
% Instruction \llvm{switch} directs the
% control flow to one of several blocks. The given expression is evaluated and
% control goes to the first branch with the matching value, or to the default
% branch (appearing first in the instruction).
Finally, instruction \llvm{ret}
ends the current function call, optionally returning a value.

The expression language is limited to names (local and global), integer literals
and selection of an element in a structure.

To conclude, we make no assumption on the existence of a library to obtain
resources managed by the operating system, such as dynamic memory allocation. In
the absence of such resource, all data must be allocated either statically, or
on the current stack frame, by using the \llvm{alloca} instruction.

\section{Translation: a global picture}
\label{sec:overview}

In addition to producing LLVM code that is faithful to the design developed in
B, our code generator is designed to satisfy the following two requirements:
\begin{enumerate}
\item Memory is allocated statically, save for the function call frame stack: no
  dynamic memory allocation library is required. Many safety critical systems
  preclude use of this programming facility.
\item The code generator features separate compilation: an internal change in a
  component should only require generating new LLVM code for that component, and
  not for the other components that depend on it. This condition is important
  for large, distributed projects.
\end{enumerate}

In Atelier-B, software development is organized in libraries of modules. A
module has a specification, called a machine. A machine may be:
\begin{description}
\item[basic] In that case its implementation is carried out outside the realm of
  the B method. Even though no LLVM code needs to be generated to implement the
  basic machines, the code generator still needs to produce LLVM declarations of
  the entities (e.g., constants and operations) that are supplied by those
  machines.
\item[developed] In that case, this module also contains a series of
  refinements, eventually reaching an implementable version, from which LLVM
  code may be generated.
\end{description}
We distinguish basic machines from developed machines in the abstract syntax by
testing the \B{impl} attribute of the \B{Machine} element: if it is $\bot$, then
the machine is basic, otherwise it is the root of the corresponding
implementation.

Amongst the several constructs of the B notation supporting modular design, the
\emph{import} relation is related to separate compilation. At the implementation
level, one module may import several instances of a machine to construct its
internal data structures. Such imported machines may be either basic, or
developed. In the latter case, the implementation of the imported machines may
in turn import other machine instances and there is no pre-established limit to
such chain of imports. Hence, the import relation between module instances forms
a tree, where the root is an implementation and the descendants of a node are
the instances imported from the module in that node.

\begin{figure}
\begin{center}
 \includegraphics[width=.8\textwidth]{B-project.pdf}
\caption{Structure of the imports relation in a B project}
\label{fig:structure}
\end{center}
\end{figure}
Figure~\ref{fig:structure} illustrates this with the structure of a B project
where an implementation \B{I0} of a specification \B{M0} imports one unnamed
instance of a basic machine \B{M1} and two instances named \B{A} and \B{B} of a
developed machine \B{M2}. Its implementation \B{I2} in turn imports two
instances \B{C} and \B{D} of a developed machine \B{M3}, implemented as \B{I3}
(which has no imports itself).  The generation of LLVM code for the project
shall produce the representation for one instance of \B{I0}, two instances of
\B{I2} and four instances of \B{I3}.

At that point, let us remark that, when generating code for a given module, we
need to consider the following two cases:
\begin{enumerate}
\item The code generation aims at producing a component, that will further be
  instantiated by importations. We call this the \emph{COMP} mode for code
  generation. This case is further split into two sub-cases:
  \begin{enumerate}
  \item The module is basic, and only declarations need to be emitted.
  \item The module is developed, and declarations and definitions are produced.
  \end{enumerate}
\item The generated code corresponds to a full project. Given the root module,
  the code generator must identify all the transitively imported module instances. We call this the \emph{PROJ} mode for code generation.
\end{enumerate}
One key issue in the code generation is how memory shall be allocated to store
the state of the modules. To satisfy our first initial requirement (no dynamic
memory allocation), this allocation shall be static, that is to say the
generator must include the code that creates the module instances when the
project starts execution. Note that this cannot be done in \emph{COMP} mode as,
at that point, we cannot assume any knowledge on the number of instances the
module being compiled will there be at run time. Consequently, the
responsibility for the static allocation of instances is assigned to the
\emph{PROJ} mode, which shall also perform the proper bindings between the
actual and formal instances in the generated code.

To encode the data space of a module, the translation produces a structure type,
named \llvm{\%M\$state\$}, such that the elements of the structure represent the
variables and the instances of imported modules. To satisfy our second initial
requirement (separate compilation), the instances cannot be part of the
structure itself, instead references to those instances (encoded as pointers)
are used. Let \llvm{\%M\$ref\$} denote the type pointer to \llvm{\%M\$state\$}.
To encode behavior of a module, each operation \B{op} is translated into the
definition of a function named \llvm{@M\$op}. The parameter list of the
functions thus produced contains one item for each input and output of the
corresponding operation. There is also one parameter of type \llvm{\%M\$ref\$},
to pass the address of the instance encoding structure associated with that
operation. Also the translation produces a special function \llvm{@M\$init\$}
responsible for executing the initialization of \B{M}. The parameters of this
function are the addresses of the instances found in the import tree of the
module (including the module itself).

Our approach consists in dividing the translation into three sections, and to
combine these sections differently in \emph{PROJ\/} and \emph{COMP\/} mode. For
a given module \B{M}, the sections are the following:
\begin{enumerate}
\item typedef: Defines an LLVM data type \llvm{\%M\$state\$} corresponding to
  the data space of the module:
  \begin{tabbing}
    \llvm{\%M\$state\$ = type \{ \ListOf{\nt{type}} \}}
  \end{tabbing}
  In case the data space of a module is empty, this section is empty and the
  type is not defined.
\item interface: Defines an LLVM type \B{M\_ref}, pointer to \B{M\_data} (if
  defined):
  \begin{tabbing}
    \llvm{\%M\$ref\$ = type \%M\$state\$*}
  \end{tabbing}
  The interface also contains declarations of one LLVM functions for each
  operation of the module.
  \begin{tabbing}
    \llvm{declare void @M\$op(\%M\$ref\$, \ListOf{\nt{type}})}
  \end{tabbing}
  If the data space of the module is not empty, a function definition
  corresponding to the initialisation is also emitted:
  \begin{tabbing}
    \llvm{declare void @M\$init\$(\%M\$ref\$, \ListOf{\nt{type}})}
  \end{tabbing}
\item implementation: Defines the functions declared in the interface:
  \begin{tabbing}
    foo \= \kill
    \llvm{define void @M\$op(\%M\$state\$* \%self\$, \ListOf{\nt{param}}) \{} \\
    \> \llvm{\ListOf{\nt{block}}} \\
    \> \llvm{exit: ret void} \\
    \llvm{\}}
  \end{tabbing}
  This section is produced only for the developed modules.
\end{enumerate}
Based on the definition of these sections, the \emph{COMP\/} translation
assembles them as follows:
\begin{center}
  \begin{tabbing}
    foo \= foo \= \kill
    \textit{for each imported module} \B{Q} \textit{generate} \\
    \> \llvm{\%Q\$state\$ = type opaque} \quad \textit{(declares type for} \B{Q} \textit{state space)} \\
    \> \textit{include the interface section of }\B{Q} \\
    \textit{include the typedef section of} \B{M} \\
    \textit{include the implementation section of} \B{M}
  \end{tabbing}
\end{center}
For a module \B{M}, the \emph{PROJ\/} translation is organized as follows:
\begin{center}
  \begin{tabbing}
    foo \= foo \= \kill
    \textit{for each transitively imported component} \B{Q} \\
    \> \textit{include the typedef section of \B{Q}} \\
    \textit{for each instance} \B{Q} \textit{imported transitively through} \B{path} \\
    \> \textit{declare a variable of type} \llvm{\%Q\$state\$}: \\
    \> \llvm{@Q$\lbrack$path$\rbrack$ = common global \%M\$state\$ zeroinitializer} \\

@@

The detailed presentation of the translation of the different elements of a B
implementation is divided into:
\begin{description}
\item[Notation] and some auxiliary definitions are presented in sec.~\ref{sec:notation}.
\item[Modularity] constructs to combine B components are described in
  sec.~\ref{sec:module}.
\item[Data] translation is briefly described sec.~\ref{sec:data}, including
  the encoding of B data types.
\item[Expressions] and conditions are codified as LLVM instruction blocks, as
  described in section~\ref{sec:expr}.
\item[Control] flow, instructions are translated to LLVM instructions as
  discussed in sec.~\ref{sec:control}, which also presents the encoding
  of B operations and initialization.
\end{description}

@@

\section{Validation}
\label{sec:validation}

Before implementing this translation, we want validate the rules that define it.
For this, we follow a pragmatic approach, based on testing and simulation of the
generated LLVM artifacts, and using the existing LLVM tool set. Consequently,
the rules described in the preceding sections have been prototyped in the Python
programming language. The input of the translation is in a XML-based format,
named BXML, that is generated by the forthcoming version of Atelier-B. For each
B component (machine or implementation), Atelier-B generates a BXML file that
may then be used as input by our prototype. The output of the translation is in
LLVM-IR format. Such output may then be processed by the LLVM tools either to
generate binary in some compatible target platform or LLVM bitcode that may then
be executed using the LLVM interpreter. We used the latter approach to validate
our translation.

The LLVM tools that are employed are:
\begin{description}
\item[llvm-as] The LLVM assembler generates LLVM bitcode, invoked, e.g., as
  follows:
\begin{verbatim}
llvm-as example-01.llvm -o example-01.bc
\end{verbatim}
\item[llvm-link] The LLVM bitcode linker, which may be invoked as follows:
\begin{verbatim}
llvm-link example-01.bc example-02.bc -o prog.bc
\end{verbatim}
\item[lli] The LLVM bitcode interpreter, which may be invoked as follows:
\begin{verbatim}
lli-mp example-01.bc
\end{verbatim}
\end{description}

We describe below the different commands required to use the LLVM framework to
animate a B implementation \B{wd} that uses an instance of a B module
\B{counter}:
\begin{verbatim}
b2llvm counter counter.llvm
b2llvm wd counter.llvm
b2llvm --mode proj wd init-wd.llvm
llvm-as counter.llvm -o counter.bc
llvm-as wd.llvm -o wd.bc
llvm-as init-wd.llvm -o init-wd.bc
llvm-as scaffold-wd.llvm -o scaffold-wd.bc
llvm-link counter.bc wd.bc init-wd.bc scaffold-wd.bc -o anim-wd.bc
lli anim-wd.bc
\end{verbatim}

\section{Conclusion and future work}
\label{sec:conclusion}

This paper presents the formal specification of a translation from a large
subset of the B implementation language to LLVM, a modern compiler internal
language.  Even though this is still a work in progress, the definition is
self-contained and has a large enough scope to be applied to B implementations
where the data belongs to basic types. This specification is a blueprint for a
code synthesis tool for B that is currently being implemented. This tool will be
distributed as an extension to Atelier-B under an open-source license.

The next step is to extend the scope to the full B implementation language. This
entails that the translation must be defined for aggregate data types and
adapted to handle some syntactic sugar. We are also planning for producing a
LLVM output with debugging information. Such output would be indeed very helpful
to provide feedback to the user when applying testing to validate the produced
code.

To prove the verification of the correctness of the translation, we would have
to define the semantics of both LLVM and B in a unified framework. One possible
starting point is Vellvm~\cite{vellvm}, a framework to reason about the
correctness of LLVM programs and transformations. We would then have to extend
the framework to include B implementation language. Another possible approach
would be to translate verification conditions from the B development artifacts
as assertinons in the generated LLVM code. The compiled program would include
checks that such assertions hold while executing.

\bibliographystyle{plain}
\bibliography{b2llvm}

\newpage
\appendix

\section{Test summary}

For each test file, indicates the B0 constructs found in the implementation.
\begin{center}
\newcommand{\no}[0]{--}
\newcommand{\yes}[0]{$\circ$}
\begin{tabular}{|l|c|c|c|c|c|c|c|c|}
\hline
Example nº & 01 & 02 & 03 & 04 & 05 & 06 & 07 & 08 \\
\hline
\hline
\multicolumn{9}{|c|}{clauses} \\
\hline
import w.o prefix & \no & \no & \no & \no & \no & \no & \yes & \no \\
import w prefix & \no & \no & \no & \no & \no & \no & \no & \yes \\
variables & \yes & \yes & \yes & \yes & \yes & \yes & \no & \yes \\
constants & \no & \yes & \no & \no & \yes & \no & \yes & \yes \\
op. input & \yes & \no & \yes & \no & \no & \no & \no & \no \\
op. output & \yes & \no & \yes & \no & \no & \no & \yes & \yes \\
\hline
\multicolumn{9}{|c|}{instructions} \\
\hline
block & \yes & \no & \yes & \yes & \no & \yes & \no & \yes \\
local var & \no & \no & \yes & \yes & \no & \yes & \yes & \yes \\
identity & \no & \no & \yes & \no & \no & \no & \no & \no \\
becomes equal & \yes & \yes & \yes & \yes & \yes & \yes & \yes & \yes \\
operation call & \no & \no & \no & \no & \no & \no & \yes & \yes \\
condition & \yes & \no & \no & \no & \yes & \no & \yes & \yes \\
case & \no & \no & \no & \no & \no & \yes & \no & \no \\
assertion & \no & \no & \no & \no & \no & \no & \no & \no \\
sequence & \yes & \yes & \yes & \yes & \no & \yes & \yes & \yes \\
while & \no & \no & \no & \yes & \no & \no & \yes & \no \\
\hline
\multicolumn{9}{|c|}{types} \\
\hline
INT & \yes & \yes & \yes & \yes & \yes & \yes & \yes & \yes \\
BOOL & \yes & \yes & \no & \no & \yes & \no & \yes & \yes \\
\hline
\multicolumn{9}{|c|}{expressions} \\
\hline
int lit & \yes & \yes & \yes & \yes & \yes & \yes & \yes & \yes \\
bool lit & \yes & \yes & \no & \yes & \yes & \no & \yes & \yes \\
vars & \yes & \yes & \yes & \yes & \no & \yes & \no & \yes \\
loc. vars & \no & \no & \yes & \yes & \no & \yes & \yes & \yes \\
const & \no & \yes & \no & \no & \yes & \no & \yes & \yes\\
succ & \no & \yes & \no & \yes & \yes & \no & \no & \no \\
pred & \no & \yes & \no & \yes & \yes & \no & \no & \no \\
+ & \yes & \no & \no & \yes & \no & \yes & \yes & \no \\
- & \yes & \no & \no & \no & \no & \no & \yes & \no \\
* & \no & \no & \no & \no & \no & \yes & \no & \no \\
\hline
\multicolumn{9}{|c|}{formulas} \\
\hline
and & \no & \no & \no & \no & \yes & \no & \no & \no \\
or & \no & \no & \no & \no & \yes & \no & \no & \no \\
not & \no & \no & \no & \no & \yes & \no & \no & \no \\
$<$ & \yes & \yes & \no & \no & \no & \no & \yes & \yes \\
$>$ & \no & \no & \no & \no & \yes & \no & \yes & \no \\
\hline
\end{tabular}
\end{center}

>>>>

\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage{amssymb}
\usepackage[fleqn]{amsmath}
\usepackage[colorlinks=false]{hyperref}
\usepackage{xspace,framed,here,keystroke}
\usepackage{graphicx}
\pagestyle{headings}
\title{Translation of B Implementations to the LLVM: Specification}
\author{David Déharbe \and Valério Medeiros Jr}
\institute{Federal University of Rio Grande do Norte \\
  Natal (Brazil)}
\date{May 2013}

@@

\section{Description of the input of the code generator}
\label{sec:b-ast}

We detail here the constructs that are considered in the compilation to the
LLVM. This description introduces the notations used in the definition of the
translation. The elements of the B language are denoted using a \B{sans serif}
font.

Figure~\ref{tab:node-attr} summarizes the structure
of the abstract syntax that is relevant in the definition of the proposed
translation and introduces identifiers used in the definition of the
translation. For instance, and for the purpose of the translation, the relevant
components of a B implementation are: \B{id}, the name identifying the
implementation within the project; \B{see}, a set of references to other modules
containing definitions used in the implementation; \B{const}, a set of constant
names, their type and their values; \B{impo}, a sequence of optionally
prefixed B machines corresponding to external modules instantiated in the
implementation; \B{var}, a set of variable names, their type and possibly
additional functional restrictions forming the rest of the state; \B{init}, a
sequence of instructions executed upon initialization of the implementation;
\B{op}, a set of operations, each being an algorithmic description of the
different functionalities provided by the implementation.

We distinguish the base module machines from developed module machines by
testing the \B{impl} attribute of the \B{Machine} element: if it is $\bot$, then
the machine is basic, otherwise it is the root of the corresponding
implementation.

@@

\begin{figure}[t]

@@

Figure~\ref{fig:llvm-grammar} presents a grammar of this subset of the LLVM.

@@

\section{Code generation modes}
\label{sec:overview}

Code generation modes are:
\begin{enumerate}
\item Code generation aims at producing a LLVM module,
  that will later be instantiated when generating the code for a project. We
  call this the \emph{COMP} mode for code generation.
\item Code generation is applied to build a full system from a project. Given
  the root module of the project, the code generator must identify all the
  transitively imported module instances. We call this the \emph{PROJ} mode for
  code generation.
\end{enumerate}

\begin{figure}[t]
\begin{center}
\newcommand{\myind}{\hspace*{2em}}
\framebox{
\begin{minipage}{.9\textwidth}
\noindent\textbf{typedef section:}
Defines an LLVM data type \llvm{\%M\$state\$} corresponding to
the data space of the module: \\
\myind \llvm{\%M\$state\$ = type \{ \ListOf{\nt{type}} \}} \\
In case the data space of a module is empty, this section is empty and the
type is not defined. \\
\\
\noindent\textbf{interface section:} 
Defines an LLVM type \B{M\_ref}, pointer to \B{M\_data} (if
defined):\\
\myind \llvm{\%M\$ref\$ = type \%M\$state\$*} \\
The interface also contains declarations of one LLVM functions for each
operation of the module. \\
\myind \llvm{declare void @M\$op(\%M\$ref\$, \ListOf{\nt{type}})} \\
If the data space of the module is not empty, a function definition
corresponding to the initialisation is also emitted: \\
\myind \llvm{declare void @M\$init\$(\%M\$ref\$, \ListOf{\nt{type}})} \\
\\
\noindent\textbf{implementation section:}
Defines the functions declared in the interface (only for developed modules): \\
\myind \llvm{define void @M\$op(\%M\$state\$* \%self\$, \ListOf{\nt{param}}) \{} \\
\myind \myind \llvm{\ListOf{\nt{block}}} \\
\myind \myind \llvm{exit: ret void} \\
\myind \llvm{\}}
\end{minipage}
}
\end{center}
\caption{Summary of the different sections and the pattern of LLVM code
  composing them.}
\label{fig:summary}
\end{figure}

Figure~\ref{fig:summary} summarizes the three sections defined in our approach
for the code generation. These sections are combined differently in
\emph{PROJ\/} and \emph{COMP\/} mode. The \emph{COMP\/} translation assembles
them as follows:
\begin{center}
  \begin{tabbing}
    foo \= foo \= \kill
    \textit{for each imported module} \B{Q} \textit{generate} \\
    \> \llvm{\%Q\$state\$ = type opaque} \quad \textit{(declares type for} \B{Q} \textit{state space)} \\
    \> \textit{include the interface section of }\B{Q} \\
    \textit{include the typedef section of} \B{M} \\
    \textit{include the implementation section of} \B{M}
  \end{tabbing}
\end{center}
For a module \B{M}, the \emph{PROJ\/} translation is organized as follows:
\begin{center}
  \begin{tabbing}
    foo \= foo \= \kill
    \textit{for each transitively imported component} \B{Q} \\
    \> \textit{include the typedef section of \B{Q}} \\
    \textit{for each instance} \B{Q} \textit{imported transitively through} \B{path} \\
    \> \textit{declare a variable of type} \llvm{\%Q\$state\$}: \\
    \> \llvm{@Q$\lbrack$path$\rbrack$ = common global \%M\$state\$ zeroinitializer} \\

@@

@@
